---
title: "JOUR472/772 lab_02 | opioid shipment analysis"
author: "Sean Mussenden"
date: "9/7/2020"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook 

This notebook will review the concepts from the in_class_02 assignment.  

* Loading packages.
* Overview of the Tidyverse suite of packages, and its opinionated approach to data analysis. 
* Programming with pre-built functions, specifically those made available with the Tidyverse.
* Loading data and getting a brief overview of what it contains. 

We will also introduce some new concepts, including

* Rearranging columns with select().  
* Sorting with arrange()
* Chaining together functions to transform data in an orderly way. 
* Debugging problems.

## How to work through it

This notebook contains text, code that I have written.  

When you see **Task**, that's an indication that you should do one of three things as instructed: 

* Run some code I've written.
* Modify some code I've written and run it.
* Write some code of your own and run it. 
* Write some text. 

In a separate video I've given you access to, I'll work through this notebook, and you should follow along the video as you work through, pausing as you need to in order to complete tasks. 

When you see **Details in Video**, that's an indication that I'll demonstrate something in the video, or explain a concept in more detail, so it will be critical to watch and listen. 

## Submission details

When you're finished all the tasks, you should save your file, commit the changes in GitHub desktop and push the changes to your repo on GitHub.com.  Then you'll submit the GitHub.com link on ELMS. 

## Packages

### About R Packages

R packages are pre-written bundles of code that provide shortcuts -- typically functions -- to help you do all kinds of things in the data programming universe -- visualize data, clean data, pull data from Twitter, scrape websites, and a whole lot more -- with less work than would be required if you wrote them on your own. 

There are thousands of free packages available for your use. 

You can load packages from directly from RStudio from the [CRAN repository](https://cran.r-project.org/web/packages/index.html). 

There are lots of packages people post on GitHub, and it's possible to load directly from GitHub.      

* [All packages are on the CRAN repository](https://cran.r-project.org/)
* [Some greatest hits](https://awesome-r.com/)
* Some of my favorites do very specific things:
  * [Tidycensus](https://github.com/walkerke/tidycensus)
  * [rTweet](https://github.com/ropensci/rtweet)
  * [Opioid data](https://cran.r-project.org/web/packages/arcos/index.html)
  * [baseballr](https://github.com/BillPetti/baseballr)
  * [spotifyr](https://github.com/charlie86/spotifyr)

### About The Tidyverse

One particularly useful collection of packages is called the Tidyverse. It makes doing data cleaning, data wrangling, data analysis and data visualization easier. 

This tutorial makes use of several Tidyverse packages to load, wrangle and analyze data. You can do everything we're doing with the base R language, but it's more of a slog.  My recommendation: if you're using R, use Tidyverse methods instead of base R whenever possible. 

When we work with R in this class, we'll generally be working with the Tidyverse, and the specific functions and methods it enables. 

That means that when we search for help solving a specific problem on the internet, it's useful to add "tidyverse r" when searching.  Here are some useful help links to bookmark for the future. 

* [Tidyverse Site](https://www.Tidyverse.org/)
* [Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
* [List of Tidvyerse packages](https://www.Tidyverse.org/packages/)
  * [Data viz with ggplot2](https://ggplot2.tidyverse.org/)
  * [Loading data with readr](https://readr.tidyverse.org/)
  * [Analyzing data with dplyr](https://dplyr.tidyverse.org/)
* [Canonical Book](https://r4ds.had.co.nz/index.html)
* [Tutorials](https://rstudio.cloud/learn/primers/)

**Details in Video**: A brief review of how to use the linked help sources to solve problems.  

### Install packages

To use a package, and the functions and methods it enables, you first have to install it. The function to install a package is `install.packages()`.  

Though there are exceptions, you typically need to install a package once -- the first time you're using a new machine. 

**Task**: Run the code block below to install the Tidyverse.  You may have already done this in class.  If you have, there's no harm in reinstalling it now.  If you run into errors, please reach out to your professor or a classmate to debug. 

```{r}
# Install the tidyverse
install.packages('tidyverse')
```

### Loading Packages

Once a package is installed on your system, you shouldn't have to install it again -- unless you want to update it to get new features.  But generally, we only need to install a package once. 
But to make use of a package, we have to load it every time we start a new session.  

Using the `library()` function is one way to do that. 

**Task**: Run the code block below to load the Tidyverse.  

```{r}
# Load the tidyverse suite of packages
library(tidyverse)
```
**Details in Video**: I'll show you what it will look like if you've installed and loaded the tidyverse correctly.  

## Loading Data

Using a function from the readr package called `read_csv()`, we're going to read in data on opioid shipments from the Washington Post's release of DEA ARCOS data.  

The data we'll load has a row for each pharmacy in Mingo County, West Virginia. It lists the total number of pills shipped between 2006 and 2014 (total_dosage_unit) and the number of shipments (total_records).

**Task**: Run the code block below to load the data.

```{r}
# Read in data for Mingo, WV
arcos_wv_mingo <- read_csv("data/arcos-wv-mingo-54059-pharmacy.csv")

```
**Details in Video**: I review in more detail what the code above means in English, a few other details about loading data in general, and some additional information about this specific data set we're loading. 

We read in the data using a function, `read_csv()`, from the readr package, which is part of the tidyverse.  We're just scratching the surface of what we can do with readr, and we'll get more advanced as the semester progresses. 

Here are some great resources for getting help with loading data using readr.

* [readr package page](https://readr.tidyverse.org/index.html).
* [details on readr functions](https://readr.tidyverse.org/reference/index.html)
* [readr cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/data-import.pdf)

**Task**: Now it's your turn.  There's a csv file in the data folder called "arcos-md-baltimore-city-24510-pharmacy.csv".  It contains the same type of information as the Mingo, WV file we read in above, but it has pill and shipment totals per pharmacy in Baltimore City.  

In the empty codeblock below, read in the Baltimore data and store it as an object called acros_md_baltimorecity

```{r}

```

**Details in Video**: I'll show you the answer, and review some common mistakes.  

## Examining data

Now that we've loaded data in, there are several ways to get a sense of the data before beginning analysis. I'll run through three common methods. 

### Option 1: glimpse
The `glimpse()` function is useful to get a sense of the data. 

**Task**: run the codeblock below to use the `glimpse()` function to get a sense of the Mingo, WV data. 
**Details in Video**: I'll go over how to make sense of the output of the glimpse function after you run it. 

```{r}

glimpse(arcos_wv_mingo)

```


**Task**: Now it's your turn. In the empty codeblock below, use the `glimpse()` function to get a sense of the Baltimore data. 

```{r}

```

### Option 2: display the data in the markdown document

By just writing the name of the object we want to explore, we can display it in our markdown document. 

**Task**: run the codeblock below to display the Mingo, WV data in the markdown document. 
**Details in Video**: I'll go over how to cycle through the data after you've printed it. 

```{r}

arcos_wv_mingo 

```

**Task**: Now it's your turn. In the empty codeblock below, display the Mingo, WV data in the markdown document. 

```{r}

```

### Option 3: open the data from the environment window

We can also click on the object name in the environment window and interact with it, resorting it to get a sense of the rows and columns. 

**Details in Video**: I'll show you how to do this.
 
## Select Data

One of the most basic things we can do a dataframe we've loaded is to trim it to only keep certain columns.  

For that, we use a select() function, which is part of dplyr. Here's the link to the [dplyr reference page](https://dplyr.tidyverse.org/index.html). 

We're also using a "pipe", which is what we call the %>% sign. It's a core part of the tidyverse, and it's what allows us to chain together functions.  In English, it means "and then.

What we're saying below is "Take the acros_wv_mingo dataframe AND THEN select only the column called buyer_name, and print it out in the markdown document"

```{r}

arcos_wv_mingo %>%
  select(buyer_name)

```
We can also store this as a new object called "arcos_wv_mingo_pharmacies", which makes it available for later use as just a list of pharmacy names. 


"Take the acros_wv_mingo dataframe AND THEN select only the column called buyer_name, and store it as a new dataframe called arcos_wv_mingo_pharmacies"

```{r}

arcos_wv_mingo_pharmacies <- arcos_wv_mingo %>%
  select(buyer_name)

```

There's lots more we can do with select, but we'll save that for this week's lab. 

For now, let's just use select to keep only a handful of columns that we want, and move others around.  So as not to overwrite the original data, let's save it as a new object called: arcos_wv_mingo_trimmed. The five columns we want to keep are: buyer_dea_no, buyer_name, buyer_city, total_dosage_unit, total_records.
 

```{r}

arcos_wv_mingo_trimmed <- arcos_wv_mingo %>%
  select(buyer_dea_no, buyer_name, buyer_city, total_dosage_unit, total_records)

```

## Sorting

By working in RStudio's spreadsheet view, we can click on a column header to re-order our spreadsheet from highest to lowest or lowest to highest, based on the values in that column.

And, of course, we can do it with code, using the arrange function, and passing in the name of a column. This puts them in order of total pills sent to each pharmacy.   

```{r}
arcos_wv_mingo_sorted <- arcos_wv_mingo_trimmed %>%
  arrange(total_dosage_unit)

```

We can also sort on the same column, highest to lowest, by wrapping the column name INSIDE of another function, called desc(), which is short for descending.  

```{r}
arcos_wv_mingo_sorted <- arcos_wv_mingo_trimmed %>%
  arrange(desc(total_dosage_unit))

```

That spreadsheet we just created serves to answer this question: which pharmacy in Mingo, W.V. got more total pills than any other pharmacy in the county between 2006 and 2014?  The answer: Strosnider. We already know that Mingo had one of the highest rates of pills per person in the country.  Could this pharmacy be a jumping off point for further reporting on a story?  Maybe. 

Let's try sorting by total_records, which is shipments.  

```{r}
arcos_wv_mingo_sorted <- arcos_wv_mingo_trimmed %>%
  arrange(desc(total_records))

```

That spreadsheet we just created serves to answer this question: which pharmacy in Mingo, W.V. got more total shipments than any other pharmacy in the county between 2006 and 2014?  The answer: Hurley Drug Company.  [Story](https://www.wvgazettemail.com/news/health/drug-firms-shipped-20-8m-pain-pills-to-wv-town-with-2-900-people/article_ef04190c-1763-5a0c-a77a-7da0ff06455b.html) 

## Putting it all together

We started by selecting columns, and then moved onto filtering.  We did this in two separate steps, but there's no reason we have to do it that way. The beauty of the Tidyverse is chaining steps together. 

The function below, in English, says: "create a new object called arcos_wv_mingo_trimmed_sorted that takes the original arcos_wv_mingo dataframe AND THEN keeps only the five columns we want AND THEN sorts the whole thing from highest to lowest on total shipments. 

```{r}
arcos_wv_mingo_trimmed_sorted <- arcos_wv_mingo %>%
  select(buyer_dea_no, buyer_name, buyer_city, total_dosage_unit, total_records) %>%
  arrange(desc(total_records))

```

## Errors

What's wrong with this code? 

```{r}
arcos_wv_mingo_trimmed <- arcos_wv_mingo %>%
  select(buer_dea_no)
```

## Exercise

There's a spreadsheet in the data folder called arcos-md-baltimore-city-24510-pharmacy.csv.  Like the mingo data we've been working with, it includes the total number of pills and total number of shipments to each pharmacy in Baltimore City between 2006 and 2014.   

Create as many codeblocks below as you need to.  Your task is to read in the data and answer these question by writing code: which pharmacy in Baltimore got the most total shipments?  Which one got the most pills?  

You'll go into breakout rooms to do this. 


