---
title: "JOUR472/772 in_class_01 | COVID reporting analysis"
author: "Sean Mussenden"
date: "8/30/2020"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook 

Imagine you're a reporter for a national news organization and your editor has given you an assignment: find a place that has been uniquely hard hit by the COVID-19 epidemic and go find a story in that community, using it as a base from which to report out a story on the epidemic.  

Where do you go?  This data notebook is designed to help you use data to narrow the list of communities, so that you can make a data driven decision about where there might be a story worth telling right now.

It's not the end product.  You're not going to publish the stats you produce here yet, though some may find their way into the story you ultimately produce.

Data is a source. If you wanted to talk to a human source, you'd talk to them on the phone and ask them questions. This notebook is your way of talking to the data, in translating your questions into a format the data understands.  

## Load libraries and settings

Behind the scenes, we're loading pre-written packages of code that other people have written to help make our analysis easier.   

```{r include=FALSE}
## Load libraries and settings
library(tidyverse)
library(lubridate)
library(janitor)
library(tidycensus)
library(maps)
library(mapview)
library(scales)
library(mapview)
library(ggthemes)
library(sf)
library(leaflet)
library(leafpop)
library(leafem)
#library(raster)
library(tigris)
library(DT)
library(moderndive)
library(kableExtra)

# Define API Key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# Turn off sci notation
options(scipen = 999)
```


## Load Data

For this analysis, we're load and cleaning data from the [U.S. Census](https://www.census.gov/) and the New York Times [repository of COVID-19 cases and deaths](https://github.com/nytimes/covid-19-data). We've reviewed the documentation and we undestand its flaws. 

```{r include=FALSE}

### Read in current data from the New York Times county-by-county COVID tracking from GitHub
# Read in county historical data on Aug 31
#covid_county <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")
#write_csv(covid_county, "covid_county_current_08_31_2020.csv") 
# Read in state historical data on Aug 31
#covid_state <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
#write_csv(covid_state, "covid_state_current_08_31_2020.csv") 
# Read in US historical data on Aug 31
#covid_us <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us.csv")
#write_csv(covid_us, "covid_us_current_08_31_2020.csv") 

### Read in stored data pulled from NYT Github on Aug 31
covid_county <- read_csv("covid_county_current_08_31_2020.csv") %>%
  filter(!state %in% c("Northern Mariana Islands","Puerto Rico","Virgin Islands", "Guam")) %>%
  filter(!county %in% c("Unknown")) %>%
  dplyr::select(fips_code = fips, county, state, date, cases, deaths)

covid_state <- read_csv("covid_state_current_08_31_2020.csv") %>%
  filter(!state %in% c("Northern Mariana Islands","Puerto Rico","Virgin Islands", "Guam"))  %>%
  dplyr::select(fips_code = fips, state, date, cases, deaths)
  
covid_us <- read_csv("covid_us_current_08_31_2020.csv") %>%
  mutate(location="US") %>%
  dplyr::select(location, everything())

### Read in population data from census
population_county <- get_acs(geography="county", variables=c("B01001_001"),geometry = FALSE, year=2018) %>%
  clean_names() %>%
  dplyr::select(fips_code=geoid,
         population=estimate) 

population_state<- get_acs(geography="state", variables=c("B01001_001"),geometry = FALSE, year=2018) %>%
  clean_names() %>%
  dplyr::select(fips_code=geoid,
         population=estimate) 

population_us <- get_acs(geography="us", variables=c("B01001_001"),geometry = FALSE, year=2018) %>%
  clean_names() %>%
  dplyr::select(fips_code=geoid,
         population=estimate) 

###Join population data to covid data
covid_county <- covid_county %>%
  left_join(population_county, by=c("fips_code")) %>%
  mutate(population = case_when(county == "New York City" ~ 8398748,
                                county == "Kansas City" ~ 491918,
                                county == "Joplin" ~ 50657,
                                TRUE~population)) %>%
  mutate(deaths_per_1000 = deaths/population*1000,
         cases_per_1000 = cases/population*1000) %>%
  dplyr::select(fips_code, county, state, date,deaths_per_1000, cases_per_1000, deaths, cases, population)

covid_state <- covid_state %>%
  left_join(population_state, by=c("fips_code")) %>%
  rename(location = state) %>%
  mutate(deaths_per_1000 = deaths/population*1000,
         cases_per_1000 = cases/population*1000) %>%
  dplyr::select(fips_code, location, date, deaths_per_1000, cases_per_1000, deaths, cases, population)

covid_us <- covid_us %>%
  mutate(fips_code = "1") %>%
  left_join(population_us, by=c("fips_code")) %>%
  mutate(deaths_per_1000 = deaths/population*1000,
         cases_per_1000 = cases/population*1000) %>%
  dplyr::select(fips_code, location, date, deaths_per_1000, cases_per_1000, deaths, cases, population)

rm(population_county, population_state, population_us)

# Join state and us data

covid_state_us <- covid_state %>%
  bind_rows(covid_us)

rm(covid_state, covid_us)

### Make a dataframe of current cases 
covid_county_current <- covid_county %>%
  filter(date == "2020-08-30")

covid_state_us_current <- covid_state_us %>%
  filter(date == "2020-08-30")

### Make a dataframe of cumulative monthly cases per capita as of 30th day of each month

covid_county_monthly_case_rate <- covid_county %>%
  filter(str_detect(date,"2020-03-30|2020-04-30|2020-05-30|2020-06-30|2020-07-30|2020-08-30")) %>%
  mutate(date = paste0("d-",date)) %>% 
  mutate(date = str_replace_all(date,"-","_")) %>%
  mutate(cases_per_1000 = cases/population*1000) %>%
  dplyr::select(county:date, cases_per_1000) %>%
  pivot_wider(names_from = date, values_from = cases_per_1000) %>%
  mutate_if(is.numeric , replace_na, replace = 0)

### Dataframe of cumulative monthly deaths per capita as of 30th day of each month
covid_county_monthly_death_rate <- covid_county %>%
  filter(str_detect(date,"2020-03-30|2020-04-30|2020-05-30|2020-06-30|2020-07-30|2020-08-30")) %>%
  mutate(date = paste0("d_",date)) %>%
   mutate(date = str_replace_all(date,"-","_")) %>%
  mutate(deaths_per_1000 = deaths/population*1000) %>%
  dplyr::select(county:date, deaths_per_1000) %>%
  pivot_wider(names_from = date, values_from = deaths_per_1000) %>%
  mutate_if(is.numeric , replace_na, replace = 0)

### Shapefiles
counties <- counties(cb = TRUE)

```

## The Interview

### Q1 State death rate
**Q:** As of August 30, what state had the highest average death rate (deaths per 1000)?

**A:** New Jersey, at 1.79 deaths per 1000.  Hawaii had the lowest. 

```{r echo=FALSE}
covid_state_us_current %>%
  dplyr::select(location, deaths_per_1000) %>%
  arrange(desc(deaths_per_1000))  %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)

  
```

### Q2 State case rate
**Q:** As of August 30, what state had the highest average case rate (cases per 1000)?

**A:** Louisiana at 31.75 cases per 1000. Hawaii had the lowest. 


```{r echo=FALSE}
covid_state_us_current %>%
  dplyr::select(location, cases_per_1000) %>%
  arrange(desc(cases_per_1000)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)

  
```

### Q3 County death rate
**Q:** As of August 30, what county had the highest death rate (deaths per 1000 people)?

**A:** Hancock, Georgia at 4.59 deaths/1000

```{r echo=FALSE}

covid_county_current %>%
  dplyr::select(county, state, deaths_per_1000) %>%
  arrange(desc(deaths_per_1000)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)

```

### Q4 County case rate
**Q:** As of August 30, what county had the highest case rate (cases per 1000 people)?

**A:** Trousdale, Tenneessee at 169 cases per 1000
```{r echo=FALSE}

covid_county_current %>%
  dplyr::select(county, state, cases_per_1000) %>%
  arrange(desc(cases_per_1000)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)


```

### Q5 Mapping deaths
**Q:** Are there any trends we can detect geographically with deaths?

**A:** Again, the south has been uniquely hard hit relative to the rest of the country.

```{r echo=FALSE}
# Join county shapes to other shapes
covid_county_population_geo <- geo_join(counties, covid_county_current, 'GEOID', 'fips_code', how = "inner")  

# Define Color Scheme
binpal <- colorBin("plasma", covid_county_population_geo$deaths_per_1000, 10, pretty = FALSE)

# Draw map

leaflet(covid_county_population_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(deaths_per_1000), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(covid_county_population_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
    addLegend("bottomleft", 
             pal = binpal, 
             values = covid_county_population_geo$deaths_per_1000,
    title = "Deaths per 100K",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) 
```

### Q6 Mapping cases
**Q:**  Are there any trends we can detect geographically with cases?

**A:** The south has been uniquely hard hit, relative to the rest of the country

```{r echo=FALSE}
# Join county shapes to other shapes
covid_county_population_geo <- geo_join(counties, covid_county_current, 'GEOID', 'fips_code', how = "inner")  

# Define Color Scheme
binpal <- colorBin("plasma", covid_county_population_geo$cases_per_1000, 10, pretty = FALSE)

# Draw map

leaflet(covid_county_population_geo) %>%
   addProviderTiles(providers$CartoDB.Positron) %>%
   #addProviderTiles(providers$Wikimedia) %>%
   addPolygons(fillColor = ~binpal(cases_per_1000), weight = 1, smoothFactor = 0.5, opacity = 0.1, fillOpacity = 0.5, color="black", popup = popupTable(covid_county_population_geo)) %>%
   setView(-95, 39.335359608681216, 4) %>%
    addLegend("bottomleft", 
             pal = binpal, 
             values = covid_county_population_geo$cases_per_1000,
    title = "Cases per 100K",
    labFormat = labelFormat(prefix = ""),
    opacity = 1
  ) 
```

### Q7 Recent increases in deaths
**Q:** In which counties have deaths gone up the most over the last month?

**A:** A lot of places in the South.

```{r echo=FALSE}

covid_county_monthly_death_rate %>%
  dplyr::select(county, state, d_2020_07_30, d_2020_08_30) %>%
  mutate(late_july_to_late_august_pct_change_death_rate = (d_2020_08_30-d_2020_07_30)/d_2020_07_30*100) %>%
  arrange(desc(late_july_to_late_august_pct_change_death_rate)) %>%
  filter(d_2020_07_30 != 0) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)

```

### Q8 Recent increases in cases
**Q:** In which counties have cases gone up the most over the last month?

**A:** A lot of places in the Midwest.

```{r echo=FALSE}

covid_county_monthly_case_rate %>%
  dplyr::select(county, state, d_2020_07_30, d_2020_08_30) %>%
  mutate(late_july_to_late_august_pct_change_case_rate = (d_2020_08_30-d_2020_07_30)/d_2020_07_30*100) %>%
  arrange(desc(late_july_to_late_august_pct_change_case_rate)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = T)

```