---
title: "JOUR472/772 in_class_10 | charts"
author: "Sean Mussenden"
date: "11/03/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook

This notebook is designed as an in-class introduction to concepts we'll delve deeper into during lab_10.  In particular, it will show:

* Exploratory data visualization.

## Install packages

There are no packages to install for this assignment.  The visualization library, ggplot2, comes installed as part of the Tidyverse.

## Load packages

Today, we're going to load one package:

* The Tidyverse. [Tidyverse Site](https://www.Tidyverse.org/)

**Task (Run Code)**: run the code below to load the tidyverse and janitor, using the library() function.

```{r}
# Load the tidyverse
library(tidyverse)


```

## Load Data

Load a dataframe of COVID case rates for Maryland counties on the 30th day of each month between March and August.  Rate is cases per 1000 population.  Source [NYT Covid GitHub repo](https://github.com/nytimes/covid-19-data).

```{r}
maryland <- read_csv("data/maryland_county_monthly_case_rate.csv")
```

Now, let's glimpse the data. 

```{r}
glimpse(maryland)
```

## Visualize data 

Our goal here is to make a simple bar chart showing the rise in the case rate in Anne Arundel county between March and August. 

### Filter data

First, let's filter the data to create a data frame with only Anne Arundel city data. 

```{r}
anne_arundel <- maryland %>%
  filter(county == "Anne Arundel")

```

### Make a chart

We'll use two functions here to make a chart, both from the ggplot2 library. First, we'll use ggplot() to initialize an empty chart.

```{r}
# Take the anne_arundel data frame 
  # Draw an empty chart with month column on the x axis and case_rate column on the y axis.  

anne_arundel %>%
  ggplot(aes(month, case_rate)) 
  
```
Now, let's display the data as a bar chart, using another function -  geom_bar() - to draw a bar chart. 

```{r}
# Take the anne_arundel data frame. 
  # Make a chart with month on the x axis and case_rate on the y axis
  # Display the data as a bar chart
anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity") 
  
```
### Continue building the chart

We can keep adding functions from the ggplot2 library to improve the chart, using a system called "the grammar of graphics". 

```{r}
# Take the anne_arundel data frame. 
  # Make a chart with month on the x axis and case_rate on the y axis
  # Display the data as a bar chart
anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity") 
  
```

-30-
