---
title: "JOUR472/772 lab_01 | Environment Setup "
author: "Sean Mussenden"
date: "9/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab objectives

This lab serves as a foundation for all the work we'll be doing this semester, helping you set up your computer to do classwork and submit completed classwork for grading. 

This lab has several objectives for you to complete, and will walk you step-by-step through completing those objectives.  
The objectives are:

* Objective 01_a: Install on your local machine the necessary software tools for this course.
* Objective 01_b: Set up a GitHub repository for classwork
* Objective 01_c: Develop an understanding of how we'll use RStudio
* Objective 01_d: Develop an understanding of how to submit lab to ELMS, and submit it

## How this lab works

You should read the text on this web page and watch the embedded videos, and follow the instructions I give as you go along.  It's very important to watch the videos, because, in addition to showing you how to do things step-by-step, the audio track also contains some key contextual information to help make sense of what you're seeing.     
## Objectives 

### Objective 01_a: Install on your local machine the necessary software tools for this course.

To complete this objective, you'll need to install on your machine the necessary software tools for this course. 

They are:

  * R, the programming language we'll be using to do data programming in this course. [Download link](https://cloud.r-project.org/)
  * R Studio, the software tool we'll be using to write R code and load, clean and analyze data. [Download link](https://rstudio.com/products/rstudio/download/#download)
  * GitHub Desktop, a software tool we'll be using to transfer code to our local machine to the GitHub.com website. [Download link](https://desktop.github.com/)

In the video below, I show you step-by-step how to do that, while offering some thoughts on what each thing does.  

Before moving on to the next objective, please watch the video in full and complete the objective by successfully installing the three items named above.      

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/zNzZ1PfUDNk")

```

### Objective 01_b: Set up a GitHub repository for classwork 

To complete this objective, you'll need to: 
  * Create an account on GitHub.com, a code repository website where we'll be storing much of the work you do in this class.  
  * Create a private repository on GitHub.com specifically for this class and give me permission to access it. 
  * Use GitHub Desktop to pull your class repository to your local machine, a process called "cloning".  
  
In the video below, I show you step-by-step how to do that, while explaining what you need to know about GitHub, as we'll use it in this class. 

Before moving on to the next objective, please watch the video in full and complete the objective by successfully doing the three objectives above.      

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/zNzZ1PfUDNk")
```

### Objective 01_c: Develop an understanding of how we'll use RStudio

To complete this objective, you'll need to:
  * Watch me tour the RStudio workspace.
  * Load a sample RMarkdown file, a file format designed for "data notebooks" that we'll use all semester. 
  * Make some minor edits to the sample RMarkdown file and run some very basic code. [Download link for the file]("https://www.youtube.com/embed/zNzZ1PfUDNk")
  
In the video below, I show you step-by-step how to do that, while explaining some fundamentals about R, RStudio and RMarkdown. 

Before moving on to the next objective, please watch the video in full and complete the objective by successfully doing the three objectives above.      

### Objective 01_d: Develop an understanding of how to submit changes to GitHub, and submit to ELMS

To complete this objective, you'll need to:
  * Save changes you've made to the RMarkdown file
  * Upload those changes from your local machine to your code repository on GitHub.com.
  * Submit your lab assignment on ELMS by sending the link to your GitHub.com repository. 
  
In the video below, I show you step-by-step how to do that, while explaining some fundamentals about GitHub

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/zNzZ1PfUDNk")
```

## Lab complete

If you have successfully submitted the link to the edited RMarkdown file in your repository on GitHub.com to the ELMS assignment for this lab, you are done!  