---
title: "JOUR472/772 in_class_05 | heat analysis, grouping and summarizing"
author: "Sean Mussenden"
date: "9/28/2020"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook 

This notebook is designed as an in-class introduction to concepts we'll delve deeper into during lab_05.  In particular, it will show:

* grouping data by one or more features
* calculating summary statistics, including counting, averages, min and max

## Install packages

There are no packages to install for this lab. 

## Load packages

Remember, we typically need to install packages only once on our computer.  But we have to load them every time we want to make use of them during an R session.  

Today, we're going to load one packages:

* The Tidyverse. [Tidyverse Site](https://www.Tidyverse.org/)

**Task (Write and Run Code)**: write the necessary code below to load the tidyverse and lubridate, using the library() function.

```{r}
# Load the tidyverse
library(tidyverse)
```

## Load Data

From the New York Times COVID-19 tracking project GitHub, dataset has cumulative totals of cases and deaths by county.  There is one row per county, with totals for that county. 

* State | character | state
* County | character | county
* FIPS | character | unique id for each county
* cases | numeric | number of cumulative confirmed COVID-19 cases on 2020-08-30
* deaths | numeric | number of cumulative confirmed COVID-19 cases on 2020-08-30

```{r}

# Read in the data.
covid_county <- read_csv("data/covid_county_2020_08_30.csv")

```

Today, we're going to **tranform** the raw data, with one record per county, through a process called grouping and summarizing, and use it to answer some questions about what's happening at the state level.  

Here's a peek at where we're headed.  

Read in the data below, which I produced ahead of time by grouping and summarizing the raw data. 

```{r}
# Read in the data 
covid_state <- read_csv("data/covid_state_2020_08_30.csv")
```

There's one record per state. The values in the "total_cases" column were produced by adding up all of the values in "cases" for each county.  The values in the "total_deaths" column were produced by adding up all of the values in "deaths" for each county.

I'll review this on Zoom, using Delaware as an example. 

For now, we're going to remove covid_state from our environment, using a new function rm(). 

```{r}
# rm() removes objects from our environment. 
rm(covid_state)
```

## Grouping and Summarizing

Today, we'll be asking a series of questions about COVID-19 cases and deaths at the level of states.  

But, as we saw, our raw data is stored at the level of counties.      

To answer questions about states, we need to do the following to our raw:

* Group all of the rows that represent a single state together.
* Calculate some summary statistic (count, average, min, max) from the values in either cases or deaths. 

For those of you who have worked in Excel or Google Sheets and created a "pivot table", this is the same process. 

And it's an important one, because this process allows us to work with data in it's more original form.  We don't have to rely on others to do the summarization for us! 


## Grouping and Summarizing | Adding

Let's answer some questions! 

Q: Which state had the fewest total covid cases on August 30?

**Task (Run Code)**: Run the codeblock below and examine the answer.  

```{r}

# Group by state, add up all cases for each state, sort from lowest to highest
# I'll review what's happening here.
least_covid_cases_state <- covid_county %>%
  group_by(state) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(total_cases)

# Display below
most_covid_cases_state

# Write the answer in a comment
# Q: Which state had the fewest total covid cases on August 30? How many cases?
# A: 
```

Q: Which state had the fewest total covid cases on August 30?

**Task (Run Code)**: Run the codeblock below and examine the answer.  

```{r}

# Group by state, add up all cases for each state, sort from highest to lowest
# I'll review what's happening here.
most_covid_cases_state <- covid_county %>%
  group_by(state) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(desc(total_cases))

# Display below
most_covid_cases_state

# Write the answer in a comment
# Q: Which state had the most total covid cases on August 30? How many cases?
# A: 
```

## Grouping and Summarizing | Calculating averages

Let's calculate the average number of cases in each county in each state, and answer this question:

Q: Which state had the highest average number of cases in a county?

To answer that, we use the mean() function.

```{r}

# Group by state, take the average, sort from highest to lowest
# I'll review what's happening here. Show delaware
averages_county_cases_by_state <- covid_county %>%
  group_by(state) %>%
  summarise(average_county_cases = mean(cases)) %>%
  arrange(desc(average_county_cases))

# Display below
averages_county_cases_by_state

# Write the answer in a comment
# Q: Which state has the most counties in our data? How many?
# A: 
```

## Grouping and Summarizing | Counting

Not a super important question, but it helps us understand something about our data.  

Q: Which state has the most counties in our data? How many? 

This time, inside of our summarize statement, instead of sum(), we have a new function, n(). sum() adds values in a specific column.  n() counts values in a specific row. 

**Task (Run Code)**: Run the codeblock below and examine the answer.

```{r}

# Group by state, count the number of rows, sort from highest to lowest
# I'll review what's happening here.
most_counties_state <- covid_county %>%
  group_by(state) %>%
  summarise(total_counties = n()) %>%
  arrange(desc(total_counties))

# Display below
most_counties_state

# Write the answer in a comment
# Q: Which state has the most counties in our data? How many?
# A: 
```

## Grouping and Summarizing | Calculating Min and Max

We can also calculate the smallest and largest number of cases, using two new functions min() and max(). In showing you this, I'll also show you how you can create more than one summarize column at a time. 

Q: Which state had the single county with the largest case count?  

**Task (Run Code)**: Run the codeblock below and examine the answer.

```{r}

# Group by state, calculate the largest and smallest case value for each county, sort from highest to lowest.  
# I'll review what's happening here. Show Delaware as example.
min_max_state <- covid_county %>%
  group_by(state) %>%
  summarise(max_case_county = max(cases),
            min_case_county = min(cases)
  ) %>%
  arrange(desc(max_case_county))

# Display below
min_max_state

# Write the answer in a comment
# Q: Which state had the single county with the largest case count?  
# A: 


```

## Breakout Room Exercise

Okay, now it's your turn to examine Michael's house and answer some questions.  

You can create one codeblock, or create multiple codeblocks. 

* Q: Which state had the highest number of deaths? Which state had the lowest?
* Q: Which state had the county with the lowest number of total deaths? What was that number?
* Q: Which state had the highest average number of deaths per county? What was that number? 

-30-


