---
title: "JOUR472/772 lab_02 | Packages, Selecting, Sorting"
author: "Sean Mussenden"
date: "9/7/2020"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab overview

This lab serves as an introduction to installing and loading packages, loading data, getting an overview of the data, selecting and sorting. 

This lab has one **objective** for you to complete, and it will walk you step-by-step through completing that objective.  

The **objectives** is:  Download the zip file linked below to your local computer, add it to your GitHub repo, and work through the data notebook (lab_02_workbook.Rmd) step-by-step while watching the video and editing it as instructed. When you are finished, push the edited file to GitHub.com, and submit the link on ELMS. 

## How this lab works

First, download [this zip file](https://github.com/smussenden/20_fall_data_journalism/raw/master/labs/lab_02/lab_02_workbook/lab_02.zip) to your desktop and open it up. [Click the link to download](https://github.com/smussenden/20_fall_data_journalism/raw/master/labs/lab_02/lab_02_workbook/lab_02.zip)  

You should then watch the video below. 

It will guide you through the process of adding the unzipped folder to your local GitHub repo, and opening the data notebook for this lab in RStudio.  

You will then work through the data notebook, making edits as instructed, while watching/listening to the video.  

When you're done, the video will show you how to commit changes to your GitHub repo, push those changes to GitHub.com and submit the URL to ELMS.  

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/MuXowdk73GM")
```

## Lab complete

If you have successfully submitted the link to the edited RMarkdown file in your repository on GitHub.com to the ELMS assignment for this lab, you are done!  
