---
title: "JOUR472/772 lab_11 | data visualization"
author: "Sean Mussenden"
date: "11/9/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook

This notebook will review the concepts introduced in the in_class_11 assignment:   
In particular, it will show:

* data visualization.

## How to work through this notebook

This notebook contains text and code that I have written.  

You should read the text.  

When you see **Task**, that's an indication that you should do one of several things as instructed:

* **Task (Run Code): ** Run some code I've written in a codeblock and examine the output, answering a question if required.
* **Task (Edit and Run Code): ** Edit some code I've written in a codeblock, run it, and examine the output, answering a question if required.
* **Task (Write and Run Code): ** Write some code of your own inside of a codeblock I've already created and run it, answering a question if required.
* **Task (Create a Codeblock, Write and Run Code):** Create a codeblock in the existing whitespace of the markdown file, write some code inside of it, run it and answer a question if required.
* **Task (Watch Video): ** Run a codeblock that contains a YouTube video, and then watch the video. In the video, I may explain a concept, or assign you to do something that is not detailed in the text.

## Submission details

When you're finished all the tasks, you should save your file in RStudio; commit the changes in GitHub desktop; and push the changes to your repo on GitHub.com.  Then you'll submit the GitHub.com link on ELMS.

## Load Packages

There are three packages you'll need to load today:

* The Tidyverse, so we can analyze data the tidyverse way. It includes the ggplot2 package, for data visualization. 
* Vembedr, so you can load the videos directly in the markdown notebook.

**Task (Run Code)**: Run the code block below to load the Tidyverse and Vembedr. Note the special instructions in comments, if you are using a new computer to run this notebook.

```{r}
# If you are on a new machine you haven't used before, you may need to install these packages before loading.  If that's the case, remove the comment # from in front of the next two lines before running the codeblock, to install the packages before loading.
# install.packages('tidyverse')
# install.packages('vembedr')

# Load the tidyverse, vembedr
library(tidyverse)
library(vembedr)

```

## Load Data

To work through this lab, we're going to load a dataframe of COVID case rates for Maryland counties on the 30th day of each month between March and August.  The source is the [NYT Covid GitHub repo](https://github.com/nytimes/covid-19-data).  

Each row is a county in Maryland. The columns are as follows:

* state 
* county 
* month 
* case_rate (cases per 1000 population).

**Task (Run Code)**: run the code below to load data.

```{r}
maryland <- read_csv("data/maryland_county_monthly_case_rate.csv") 
```

## Examine data

Let's glimpse our data to get a sense of it, and display it in the markdown document.

**Task (Run Code)**: run the code below to glimpse the data.
```{r}
# glimpse the data
glimpse(maryland)

# view it
maryland



```

## Filter data 

First, let's filter the data to create a data frame with only Anne Arundel county data.

**Task (Run Code)**: run the code below to filter the data.

```{r}
anne_arundel <- maryland %>%
  filter(county == "Anne Arundel")

```

## Make a basic bar chart

We're going to make a bar chart showing the increase in the case rate in Anne Arundel County between March and August.  

We'll build this up progressively, adding features to the chart as we go. 

First, let's create the basic chart. 

**Task (Run Code)**: run the code below to visualize the data as a bar chart.

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity")


```

**Task (Watch Video)**: Watch video below, and I'll walk you through what happened. 

```{r}
# Video A
# Click play to run the codeblock and the video will display in the markdown document.
embed_youtube("ymaNpmN-W3E", width = "100%", height="480", frameborder = 0,
allowfullscreen = FALSE, query = NULL)

# If the video doesn't work here, load this URL in your browser:
# https://youtu.be/ymaNpmN-W3E

```

## Style the chart 

Let's make the chart a bit more presentable.  

First, let's change the color to a nice "steel blue" by modifying the geom_bar() function

**Task (Run Code)**: run the code below to change the color to steel blue. 

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity", fill="steelblue")

```

Now let's change the X and Y axis labels, changing them from the default to something a bit more presentable. For this, we'll use a function called labs() -- which is short for labels. 

**Task (Run Code)**: run the code below to change the X and Y axis labels.

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity", fill="steelblue") +
  labs(
    x="Month",
    y="Cases per 1,000 people"
  )

```

Now let's add a chart title and a source caption, to help people who see  the chart understand what they're supposed to take away. We'll add more properties to our labs() function.

**Task (Run Code)**: run the code below to add a title and caption.

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity", fill="steelblue") +
  labs(
    x="Month",
    y="Cases per 1,000 people",
    title="In Anne Arundel, steady increase in COVID cases",
    caption="Source: Analysis of New York Times COVID data"
  )

```

Now let's add some labels over each bar, to help people see specific values for each bar. For that, we'll add a new function, geom_text(), which allow us to visualize text values on the chart. 

**Task (Run Code)**: run the code below to change the X and Y axis labels.

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity", fill="steelblue") +
  labs(
    x="Month",
    y="Cases per 1,000 people",
    title="In Anne Arundel, steady increase in COVID cases",
    caption="Source: Analysis of New York Times COVID data"
  ) +
  geom_text(aes(label=case_rate))

``` 

The labels are a little unwieldy.  First, let's round the values to two decimal places by using the round() function inside our geom_text() function.

**Task (Run Code)**: run the code below to change the X and Y axis labels.

```{r}

anne_arundel %>%
  ggplot(aes(month, case_rate)) +
  geom_bar(stat="identity", fill="steelblue") +
  labs(
    x="Month",
    y="Cases per 1,000 people",
    title="In Anne Arundel, steady increase in COVID cases",
    caption="Source: Analysis of New York Times COVID data"
  ) +
  geom_text(aes(label=round(case_rate,0)))

``` 


**Task (Watch Video)**: Watch video below, and I'll walk you through some of the previous steps. 

```{r}
# Video B
# Click play to run the codeblock and the video will display in the markdown document.
embed_youtube("1EWVCIcX6no", width = "100%", height="480", frameborder = 0,
allowfullscreen = FALSE, query = NULL)

# If the video doesn't work here, load this URL in your browser:
# https://youtu.be/1EWVCIcX6no

```

## On your own

Now, it's your turn.  

Create a code block below, and create a bar chart showing the increase in case rate between MAY and AUGUST in Baltimore city. Style the chart similarly to the chart we created for Anne Arundel county.  

**Task (Create a Codeblock, Write and Run Code):** Create the chart below. 

```{r}

```

## You're done

When you're finished all the tasks, you should save your file in RStudio; commit the changes in GitHub desktop; and push the changes to your repo on GitHub.com.  Then you'll submit the GitHub.com link on ELMS.
