---
title: "JOUR472/772 in_class_08 | cleaning demonstration"
author: "Sean Mussenden"
date: "10/20/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook

This notebook is designed as an in-class introduction to concepts we'll delve deeper into during lab_08.  In particular, it will show:

* some common data cleaning tasks.

## Install packages

There is one package to install for this assignment:
* [Janitor](https://github.com/sfirke/janitor)

## Load packages
  
Today, we're going to load one package:

* The Tidyverse. [Tidyverse Site](https://www.Tidyverse.org/)

**Task (Run Code)**: run the code below to load the tidyverse and janitor, using the library() function.

```{r}
# Load the tidyverse
library(tidyverse)

# Load janitor
# If you need to install it, run this code
# install.packages('janitor')
library(janitor)

```

## Load Data

We're going to load a "dirty" example of our county-level COVID data for Maryland. The sample is from the New York Times COVID-19 tracking project GitHub, with cumulative totals of deaths by county. 

We're filtering the data as we load it, to only keep data for Maryland for each dataframe. 

**Task (Run Code)**: run the code below to load data.

```{r}

# Read in dirty COVID data as an object called county_covid_deaths_dirty

county_covid_deaths_dirty <- read_csv("data/county_covid_deaths_dirty.csv")

```
## Examine data

Let's glimpse our data.

```{r}

glimpse(county_covid_deaths_dirty)

```

And let's examine it as a dataframe.

```{r}
county_covid_deaths_dirty
```

What flaws jump out at you? 

## Questions

Let's try asking some questions.

Q: How many deaths are there in the county with FIPS code 24003 -- Anne Arundel?

```{r}
county_covid_deaths_dirty %>%
  filter(FIPS code !, == 24003)
```

Q: How many counties are in Maryland?

```{r}
county_covid_deaths_dirty %>%
  group_by(state) %>%
  count()
```

Q: How many total deaths are there in Maryland?

```{r}
county_covid_deaths_dirty %>%
  summarize(total_deaths = sum(deaths))

```

## Flaws

There are several flaws that became apparent when we tried to answer questions. 

* Multiple variations of Maryland in state column. 
* Missing value in deaths column. 
* Number written out as text in deaths column. 
* Deaths column stored as "character" instead of number.
* Funky column name that contains a comma for fips code. 

## Cleaning

### Clean up column headers

First, let's use the clean_names() function from the janitor package to standardize the column names, making them lowercase and removing special characters. 

Note: as we make changes, we're going to save our changes in a new dataframe, called covid_county_deaths. 

```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names()

county_covid_deaths

```

Let's try answering a question again.

Q: How many deaths are there in the county with FIPS code 24003 -- Anne Arundel?

```{r}
county_covid_deaths %>%
  filter(fips_code == 24003)

# A: 
```

### Fix multiple values in state column

We can use a function -- tolower() -- to make every value in the state column lowercase.  

That will fix one of the problems,  the all caps version of MARYLAND in the first row. 


```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names() %>%
  mutate(state = tolower(state))

county_covid_deaths

```

Let's try answering this question again.

Q: How many counties are in Maryland?

```{r}
county_covid_deaths %>%
  group_by(state) %>%
  count()
```

We're close, but we haven't fixed the problem in the fifth row. The state value for Calvert county is still "mary land".

To fix that, we can use a function called case_when() to edit ONLY the state value for Calvert, and keep all the rest the same.

```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names() %>%
  mutate(state = tolower(state)) %>%
  mutate(state = case_when(
    county == "Calvert" ~ "maryland",
    TRUE ~ state
  ))

county_covid_deaths

```

Let's try answering this question again.

Q: How many counties are in Maryland?

```{r}
county_covid_deaths %>%
  group_by(state) %>%
  count()
```

### Fix the deaths column 

The deaths column has several problems.  Let's fix the first one. For Caroline county, it says "three" instead of the number 3.  We can again use case_when().

```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names() %>%
  mutate(state = tolower(state)) %>%
  mutate(state = case_when(
    county == "Calvert" ~ "maryland",
    TRUE ~ state
  )) %>%
  mutate(deaths = case_when(
    deaths == "three" ~ "3",
    TRUE ~ deaths
  )) 

county_covid_deaths
```

Let's try answering our question again.

Q: How many total deaths are there in Maryland?

```{r}

county_covid_deaths %>%
  summarize(total_deaths = sum(deaths))

```

We get an error, because we can't summarize a column with the values stored as "character", instead of numbers. If we want to do math, we need to make it a numeric column. Let's fix that with the function as.numeric(). 

```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names() %>%
  mutate(state = tolower(state)) %>%
  mutate(state = case_when(
    county == "Calvert" ~ "maryland",
    TRUE ~ state
  )) %>%
  mutate(deaths = case_when(
    deaths == "three" ~ "3",
    TRUE ~ deaths
  )) %>%
  mutate(deaths = as.numeric(deaths))

county_covid_deaths
```
Let's try answering our question again.

Q: How many total deaths are there in Maryland?

```{r}

county_covid_deaths %>%
  summarize(total_deaths = sum(deaths))

```

We get NA for total deaths, which clearly isn't the right answer. 

We are getting this error because the data has NA for deaths for Garrett county.

After checking with the original source of the data, we determine that if it says NA, that means 0. 

So we can safely make the change. To convert the NA value to 0, we can use a handy function called replace_na()

```{r}
county_covid_deaths <- county_covid_deaths_dirty %>%
  clean_names() %>%
  mutate(state = tolower(state)) %>%
  mutate(state = case_when(
    county == "Calvert" ~ "maryland",
    TRUE ~ state
  )) %>%
  mutate(deaths = case_when(
    deaths == "three" ~ "3",
    TRUE ~ deaths
  )) %>%
  mutate(deaths = as.numeric(deaths)) %>%
  mutate(deaths = replace_na(deaths, 0))

county_covid_deaths
```
Let's try answering our question again.

Q: How many total deaths are there in Maryland?

```{r}

county_covid_deaths %>%
  summarize(total_deaths = sum(deaths))

```

## Takeaways

* Real data in the wild is rarely perfect.  Expect to spend a lot of time cleaning data. 
* Always examine your data for flaws. ALWAYS assume there's a flaw that is preventing you from getting the right answer. 
* Cleaning is an iterative process.  We can fix some problems from the jump. But more common is to discover something as you're trying to answer a question.  At the end of my process, I usually take all of the cleaning and put it in its own codeblock after I load data.  Load > Clean > Analyze.    

-30-