---
title: "JOUR472/772 in_class_08 | one-to-many joins demonstration"
author: "Sean Mussenden"
date: "10/20/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook

This notebook is designed to demonstrate one-to-many joins. 

## Install packages

There are no packages to install for this assignment.

## Load packages
  
Today, we're going to load one package:

* The Tidyverse. [Tidyverse Site](https://www.Tidyverse.org/)

**Task (Run Code)**: run the code below to load the tidyverse, using the library() function.

```{r}
# Load the tidyverse
library(tidyverse)

```

## Load Data

We're going to load two very small dataframes today to examine one-to-many joins.

These are samples from the New York Times COVID-19 tracking project GitHub, with cumulative totals of deaths by county. We also have population data by the state from the Census.

We're filtering the data as we load it, to only keep data for Maryland for each dataframe. 

**Task (Run Code)**: run the code below to load data.

```{r}

# Read in COVID data as an object called covid_state_deaths
# Filter to keep data only for Maryland and drop cases column
county_covid_deaths <- read_csv("data/covid_county_2020_08_30.csv") %>%
  filter(state %in% c("Maryland")) %>%
  select(-cases)

# Read in state population data as an object called state_population
# Filter to keep data only for Maryland, rename population column state_population
state_population <- read_csv("data/state_population.csv") %>%
  filter(state %in% c("Maryland")) %>%
  rename(state_population = population)

# display state_covid_deaths
county_covid_deaths

# display state_population
state_population

```

## One to many joins

When a value we're joining is duplicated in one of our tables, it's important to understand what happens. 

In our COVID table, we have one value for each Maryland county.  That means the value for state -- Maryland -- repeats 24 times.

```{r}

county_covid_deaths

```

In our population table, we have one value for Maryland in state. 

```{r}

state_population

```

When we join them together, the population data is joined 24 discrete times, even though it exists only once in the population table. 

```{r}
inner_covid_population <- county_covid_deaths %>%
  inner_join(state_population)

inner_covid_population

```
The type of join we do doesn't affect this repetition. 

For example, this repetition occurs when we do a left join.

```{r}
left_covid_population <- county_covid_deaths %>%
  left_join(state_population)

left_covid_population

```
It also occurs when we do a right join.

```{r}
right_covid_population <- county_covid_deaths %>%
  right_join(state_population)

right_covid_population

```
How can this go wrong? This was a simple example.  

If you have duplicates in both of your join tables, and aren't aware of that fact, you could end up with an erroneous result if you try to count.  

Let's read in a version of our state population table has two rows for Maryland -- one for population in 2010 and one for population to 2020. 

```{r}

state_population_duplicates <- read_csv("data/state_population_duplicates.csv")
  
state_population_duplicates

```

Now let's inner_join it to our deaths table. 

```{r}
inner_covid_population <- county_covid_deaths %>%
  inner_join(state_population_duplicates)

inner_covid_population

```
We get 48 records, not 24.  Note that for each county, we have two population values -- one for 2020 and one for 2010.  

Is this a problem.  On its face, not really.  If we tried to use this table to, say, count the number of counties in Maryland, we'd end up with 48, not 24.

```{r}
inner_covid_population %>%
  count()

```

Or if we try to add up all the deaths, we'll get a number that's double the real number. 

```{r}
inner_covid_population %>%
  summarise(total_deaths = sum(deaths))

```
-30-