---
title: "JOUR472/772 in_class_07 | more joins"
author: "Sean Mussenden"
date: "10/06/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this notebook

This notebook is designed as an in-class introduction to concepts we'll delve deeper into during lab_07.  In particular, it will show:

* a review of joins basics from prior lab.
* some additional information about joins and some new joins.

## Install packages

There are no packages to install for this assignment.

## Load packages

Remember, we typically need to install packages only once on our computer.  But we have to load them every time we want to make use of them during an R session.  

Today, we're going to load one package:

* The Tidyverse. [Tidyverse Site](https://www.Tidyverse.org/)

**Task (Run Code)**: run the code below to load the tidyverse, using the library() function.

```{r}
# Load the tidyverse
library(tidyverse)

```

## Load Data

We're going to load two very small dataframes today to reinforce our understanding of how joins work.

These are samples of dataframe from the New York Times COVID-19 tracking project GitHub, with cumulative totals of deaths by state. We also have population data from the Census.

We're filtering the data as we load it, to only keep data for four states for each dataframe.  Three of the states are the same -- Maryland, Virginia and Delaware.  The deaths table has DC, but not Pennsylvania.  The population table has Pennsylvania, but not DC.

**Task (Run Code)**: run the code below to load data.

```{r}

# Read in COVID data as an object called covid_state_deaths
# Filter to keep data only for Maryland, Virgina, Delaware, DC
state_covid_deaths <- read_csv("data/covid_state_2020_08_30.csv") %>%
  filter(state %in% c("Maryland", "Virginia","Delaware","District of Columbia")) %>%
  select(-total_cases)

# Read in state population data as an object called state_population
# Filter to keep data only for Maryland, Virgina, Delaware, Pennsylvania
state_population <- read_csv("data/state_population.csv") %>%
  filter(state %in% c("Maryland", "Virginia","Delaware","Pennsylvania"))

# display state_covid_deaths
state_covid_deaths

# display state_population
state_population

```

## Reviewing Joins

### Inner Join

Let's "inner join" state_covid_deaths to state_population and examine which records are returned. 

```{r}
inner_state_covid_population <- state_covid_deaths %>%
  inner_join(state_population)

inner_state_covid_population

# Q: which states are returned by this join?
# A: 
```


### Left Join

Let's "left join" state_covid_deaths to state_population and examine which records are returned. 

```{r}

left_state_covid_population <- state_covid_deaths %>%
  left_join(state_population)

left_state_covid_population

# Q: which states are returned by this join?
# A: 
```
### Right Join

Let's "right join" state_covid_deaths to state_population and examine which records are returned. 

```{r}

right_state_covid_population <- state_covid_deaths %>%
  right_join(state_population)

right_state_covid_population

# Q: which states are returned by this join?
# A: 
```
### Left Join -- Switch order of data frames

Let's "left join" state_population to state_covid_deaths and examine which records are returned. 

```{r}

left_state_population_covid <- state_population %>%
  left_join(state_covid_deaths)

left_state_population_covid

# Q: which states are returned by this join?
# A: 
```
## New Joins

### Full join

A "full join" returns all rows from both tables.  If there's a matching record in each table, it links them up.  Otherwise, it leaves the corresponding rows blank. 

Let's "full join" state_covid_deaths to state_population and examine which records are returned. 

```{r}

full_state_covid_population <- state_covid_deaths %>%
  full_join(state_population)

full_state_covid_population

# Q: which states are returned by this join?
# A: 
```

### Anti Join

An "anti join" helps us determine what records in a table do NOT have a match in another table.  

Let's "anti join" state_covid_deaths to state_population and examine which records are returned. Any record that is returned will exist in our state_covid_deaths table but NOT have a match in state_population. 

```{r}

anti_state_population_covid <- state_covid_deaths %>%
  anti_join(state_population)

anti_state_population_covid

# Q: which states are returned by this join?
# A: 
```

Now let's do the anti_join the other direction. Any record that's returned will exist in our state population table but NOT have a match in state_covid_deaths. 

```{r}

anti_state_covid_population <- state_population %>%
  anti_join(state_covid_deaths)

anti_state_covid_population

# Q: which states are returned by this join?
# A: 
```

-30-