---
title: "JOUR472/772 lab_01 tutorial | Environment setup"
output: learnr::tutorial
runtime: shiny
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Lab overview
a
This lab serves as a foundation for all the work we'll be doing this semester. It will help you set up your computer to do classwork by installing the necessary software tools.  It will give you an overview of how to use those tools.  And it will show you how to submit completed labwork for grading. 

This lab describes several **objectives** for you to complete, and it will walk you step-by-step through completing those objectives.  

The objectives are:

* **Objective 01_a**: Install on your local machine the necessary software tools for this course.
* **Objective 01_b**: Set up a GitHub repository for classwork.
* **Objective 01_c**: Develop an understanding of the basics of RStudio, and write code and text in RMarkdown files.
* **Objective 01_d**: Develop an understanding of how to submit changes you've made locally to your repo on GitHub.com, and submit to ELMS.


## Objective 01_a 

### Install on your local machine the necessary software tools for this course

To complete this objective, you'll need to install on your machine the necessary software tools for this course. 

They are:

  * R, the programming language we'll be using to do data programming in this course. [Download link](https://cloud.r-project.org/). [About R](https://www.r-project.org/about.html)
  * R Studio, the software tool we'll be using to write R code and load, clean and analyze data. [Download link](https://rstudio.com/products/rstudio/download/#download). [Brief explainer video](https://rstudio.com/products/rstudio/)
  * GitHub Desktop, a software tool we'll be using to transfer code to our local machine to the GitHub.com website. [Download link](https://desktop.github.com/)

In the video below, I show you step-by-step how to do that, while offering some thoughts on what each thing does.  

**For windows users**: in the video, I show how to install the R language for a Mac.  To install for Windows, go to [https://cloud.r-project.org/](https://cloud.r-project.org/), and click Download R for Windows. On the next page, click the first link, "base". On the next page, click the top link to download for Windows.

Before moving on to the next objective, please watch the video in full and complete the objective by successfully installing the three things listed above.     

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/ulKZkmcMiHQ")
```

```{r objective_01_a_check, echo = FALSE}
question_radio(
  "Have you successfully installed R, RStudio and GitHub Desktop?",
  answer("yes", message = 'Great! You are ready to move on', correct = TRUE),
  answer("no", message = 'Before moving on, make sure you have all three installed.  Having problems? Send the professor a note on ELMS.'),
   allow_retry = TRUE,
   correct = "",
   incorrect = ""
)
```

## Objective 01_b

### Set up a GitHub repository for classwork 

To complete this objective, you'll need to: 

  * Create an account on [GitHub.com](https://github.com), a code repository website where we'll be storing much of the work you do in this class. You should also  sign up for [GitHub student developer pack](https://education.github.com/students), which gives you free access to several features that you would otherwise have to pay for.   
  * Create a private repository (or public if you wish) on GitHub.com specifically for this class and give me permission to access it. 
  * Use GitHub Desktop to pull your class repository to your local machine, a process called "cloning."  
  
In the video below, I show you step-by-step how to do each of those things, while explaining what you need to know about GitHub as we'll use it in this class. 

Before moving on to the next objective, please watch the video in full and complete the objective by successfully doing the three things listed above.      
```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/je0rdg81p7s")
```

```{r objective_01_b_check, echo = FALSE}
question_radio(
  "Have you successfully created an account on GitHub.com, created a repo for this class and given the professor permission to access it, and used GitHub desktop to clone it to your local machine?",
  answer("yes", message = 'Great! You are ready to move on', correct = TRUE),
  answer("no", message = 'Before moving on, make sure you have completed all the steps.  Having problems? Send the professor a note on ELMS.'),
   allow_retry = TRUE,
   correct = "",
   incorrect = ""
)

random_praise()
```

## Objective 01_c 
### Develop an understanding of the basics of RStudio, and write code and text in RMarkdown files

To complete this objective, you'll need to:

  * Watch me describe the RStudio workspace.
  * Load a sample RMarkdown file, a file format tailor made for the "data notebooks" that we'll use all semester. [Right click to download the file lab_01_workbook](https://raw.githubusercontent.com/smussenden/20_fall_data_journalism/master/labs/lab_01/lab_01_workbook/lab_01_workbook.Rmd)
  * Make some minor edits to the sample RMarkdown file I've provided and run some very basic code. 
  
In the video below, I show you step-by-step how to do each of those things, while explaining some fundamentals about R, RStudio and RMarkdown. 

Before moving on to the next objective, please watch the video in full and complete the objective by successfully doing the three things listed above.      

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/SzqJLGEoIyY")
```

## Objective 01_d 
### Develop an understanding of how to submit changes you've made locally to your repo on GitHub.com, and submit to ELMS

To complete this objective, you'll need to:

  * Save changes you've made to the RMarkdown file in RStudio. 
  * Upload the changes made to your local repository to your code repository on GitHub.com.
  * Submit your lab assignment on ELMS by pasting in the url to your RMarkdown file in your GitHub.com repository. 
  
In the video below, I show you step-by-step how to do that, while explaining some fundamentals about GitHub.

```{r echo=FALSE}
knitr::include_url("https://www.youtube.com/embed/Ds4idIBjVvM")
```

## Lab complete

If you have successfully submitted the link to the edited RMarkdown file in your repository on GitHub.com to the ELMS assignment for this lab, you are done!  